---
title: "R Notebook"
output: html_notebook
---

# Project 2: Seasonal Dataset

1. No trend in present

  - Constant mean

2. Seasonality is present

  - Seasonal decomposition
  - Seasonal ARIMA
  - Exponential smoothing
  
3. Presence of negative values

  - Multiplicative exponential smoothing models are excluded
  
4. Stable amplitude (up and down movement)

## Getting data

```{r libraries}
library(readxl)
library(forecast)
library(ggplot2)
```
```{r importing dataset}
# for time series, no row or column IDs are included and data needs to be sorted from start to end e.g. from 2000 to 2008
# we need to work with vector, not dataframe
mydata=scan()

# -0.31	0.41	0.51	-0.2	0.61	0.2	0.61	-0.3	-0.1	-0.2	-0.51	0.41
# -0.51	0.61	-0.2	0.1	-0.1	0.3	0	0.2	-0.3	0	-0.1	0.81
# -0.6	0.4	0.5	0.1	-0.1	0	0.2	0.1	-0.1	0.1	0.1	0.6
# -0.2	0.6	0.59	0	0	0.1	0.2	0.1	0.2	0	0.2	0.19
# -0.1	0.68	0.58	-0.19	0	-0.19	0.39	0.38	0.1	0	0.1	0.29
# -0.48	0.57	0.48	-0.47	0.38	0.09	0.47	0	0	-0.19	0.19	0.38
# -0.56	0.47	0.28	-0.19	-0.09	0.28	0.28	0	0	-0.28	0	0
# -1.03	0.85	0.47	0	0.09	-0.09	0.19	0	-0.19	0	0.09	-0.09
# -0.84	0.38	0.75	-0.37	0.28	0.09	0.28	0	0.09	0.19	0.09	0.74
# -0.64	0.65	0.18	0	-0.18	0.18	0.37	0.09	0.09	0		
```

```{r plot the data before converting}
plot.ts(mydata)
```

```{r convert to timeseries}
germaninfl = ts(mydata, start=2008,frequency=12)
# Example: ts(mydata, start = c(2008,3), frequency =12); seasonal dataset needs frequency
```

```{r plot timeseries}
plot(germaninfl)

```

Project Outline:

- Seasonal Decomposition
- Exponential Smoothing
- Seasonal ARIMA
- Cross Validation

# Seasonal Decomposition

An old analytical method

Dividing the data into its components:

- Trend
- Seasonality
- Remainder / White noise

Models:

- Additive
- Multiplicative

If seasonal components stays constant over several cycles, it is best to opt for additive decomposition.

Model is easy to use and is simple. Good starting point for further data exploration.

Disadvatages:

- First few values are NA (because it uses moving averages)
- Slow to catch fast rises
- Adopts a constant seasonal component

** decompose method**

```{r seasonal decomposition with decompose}
decompose(germaninfl)
```

```{r plot decompose}
plot(decompose(germaninfl))
```

First plot on top shows the original dataset. Then we get trend component, next we see how seasonal part looks and the rest is white noice.

It shows that there is no trend in this dataset. It goes up and down.

We can see clearly seasonal component with peaks in Febrary and end of the year.

**Another method:**

X11, Seats or STL

- No NA Values
- The seasona part can be adjusted over time

Using stl, it is robust towards outliers, suited towards an additive model (for multiplicative models data transformation is needed), seasonal and trend cycles may change over time

Plotting seasonal decomposition with stl:
```{r seasonal decomposition with stl}
plot(stl(germaninfl, s.window=7))
# how many seasonal cycles are needed to calculate changes
```

Seasonal decomposition adn forecasting with stl:
```{r seasonal decomposition and forecasting with stl}
plot(stlf(germaninfl, method="ets"))
# ets = method for forecasting part' automated exponential smoothing method; alternatively arima, naive method or R.W. drift can be used
```

Comparison with standard ets forecast:
```{r comparison with a standard ets forecast}
plot(forecast(ets(germaninfl), h=24))
```

Looking at both methods, peaks are higher with first method.

First model is ETS(A,N,N) which means additive season, no tren and no error or remainder.

Second model is ETS(A,N,A) which is additive seasonality, no trend but additive error.

** Autoplot model:**

```{r using autoplot ets}
# library(ggplot2)
autoplot(stlf(germaninfl, method="ets"))
```

# Seasonal ARIMA

Seasonal ARIMA models have two sets of parameters: A regular set and a sexond set for the seasonal part

ARIMA (p,d,q) (P,D,Q) [m]

p,d,q => Non-seasonal part
P,D,Q => Seasonal part
m => Frequency

```{r auto arima}
auto.arima(germaninfl, stepwise = T, approximation = F, trace = T)
# ARIMA (1,0,2)(0,1,1)[12]
# for non-seasonal part, we have one order of autocorrelation, no differencing and 2 orders of moving averages; the seasonal part test one step of differencing and one order of moving average 
# we have 12 observations per interval (year)

# the corresponding coefficients are:
# ARIMA (1,0,2) => autoregressive coefficient 1 is -0.79, the moving average 1 is 0.7786, moving average 2 is 0.2115

# (0,1,1) => seasonal moving average 1(last one) is -0.76
```

```{r getting object}
germaninflarima = auto.arima(germaninfl, stepwise = T, approximation = F, trace = T)
```
```{r plot arima model}
forec = forecast(germaninflarima)
plot(forec)
```

It is good to compare Seasonal ARIMA to ETS Model or Holt-Winters Model.


